---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r reading, echo = TRUE}
library(ggplot2)

activity <- read.csv(unz("activity.zip", "activity.csv"))
activity <- transform(activity, date = as.Date(date, "%Y-%m-%d"))
activity.woNA <- activity[!is.na(activity$steps), ]
```  

## What is mean total number of steps taken per day?
```{r totals, echo = TRUE}
total.steps <- as.vector(tapply(activity.woNA$steps, activity.woNA$date, sum))
total.steps.df <- data.frame(total.steps)
mean.ts <- mean(total.steps)
median.ts <- median(total.steps)

ts.range = max(total.steps) - min(total.steps)

ggplot(data = total.steps.df, aes(total.steps)) + 
    geom_histogram(binwidth = ts.range/10, col = "blue", aes(fill=..count..)) + 
    geom_vline(aes(xintercept = median.ts), color = "green", size = 1)
```  

Mean of the total steps: `r mean.ts`  
Median of the total steps: `r median.ts`    

## What is the average daily activity pattern?
```{r averages}
avg.steps <- tapply(activity.woNA$steps, activity.woNA$interval, mean)
avg.steps.df <- data.frame(interval = as.numeric(names(avg.steps)), 
                           average.steps = as.vector(avg.steps))
max.avg.steps.index <- which.max(avg.steps.df$average.steps)
max.avg.steps.interval <- avg.steps.df[max.avg.steps.index, "interval"]
max.avg.steps.value <- avg.steps.df[max.avg.steps.index, "average.steps"]

ggplot(data = avg.steps.df, aes(interval, average.steps, group = 1)) +
    geom_line(aes(color = average.steps)) + 
    geom_point(data = NULL, 
               aes(max.avg.steps.interval, max.avg.steps.value), 
               color = "red", 
               size = 2)
```

Interval with the maximum number of steps on average across all days: `r max.avg.steps.interval`  
The value in that interval: `r max.avg.steps.value`  

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
